---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```

# GenomeScaleEmbeddings

## Load the package

```{r}
library(GenomeScaleEmbeddings)
library(patchwork)
```

## 1. Peek into remote parquet files

```{r}
# Use OpenRemoteParquetView to inspect the first few rows
OpenRemoteParquetView()
```

## 2. Copy remote parquet files into local DuckDB (overwrite = FALSE recommended for speed)

```{r}
CopyParquetToDuckDB(db_path = "local_embeddings.duckdb", overwrite = FALSE)
file.info("local_embeddings.duckdb")$size
```

## 3. Write embeddings to houba mmatrix (overwrite = FALSE recommended for speed)

```{r}
houba <- writeEmbeddingsHoubaFromDuckDB(dbPath = "local_embeddings.duckdb", overwrite = FALSE)
file.info("local_embeddings.houba")$size
```

## 4. Run PCA on houba mmatrix

```{r}
system.time(
  pca_res <- houbaPCA("local_embeddings.houba")
)
```

## 5. Get PCA scores

```{r}
pc_scores <- getPcaScores(pca_res)
object.size(pc_scores)
```

## 6. Plot PCA dimensions (PC1 vs PC2, colored by chromosome)

```{r}
plotPcaDims(pc_scores, houba$info, annotation_col = "chrom", dim1 = 1, dim2 = 2)
```

## 7. Print spatial correlations (PC1 vs genomic position, by chromosome)

```{r}
library(knitr)
cor_table <- data.frame(
  Chromosome = unique(houba$info$chrom),
  Correlation = sapply(unique(houba$info$chrom), function(chr) {
    idx <- houba$info$chrom == chr
    cor(pc_scores[idx, 1], houba$info$pos[idx], use = "complete.obs")
  })
)
kable(cor_table, caption = "Correlation between PC1 and Genomic Position by Chromosome")
```

---

- All file sizes are displayed after each step for transparency.
- Remote parquet files are previewed before local processing.
- For large datasets, consider subsampling for faster plotting.
- Use `overwrite = FALSE` for DuckDB and houba steps to avoid slow download and reuse existing files.
