---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  fig.path = "docs/README-"
)
```

# GenomeScaleEmbeddings

## Load the package

```{r}
library(GenomeScaleEmbeddings)
library(knitr)
library(httr2)
library(jsonlite)
library(dplyr)
library(ggplot2)
```

## 1. Peek into remote parquet files

```{r}
# Use OpenRemoteParquetView to inspect the first few rows
OpenRemoteParquetView()
```

## 2. Copy remote parquet files into local DuckDB

```{r}
system.time(

CopyParquetToDuckDB(db_path = "local_embeddings.duckdb", overwrite = FALSE)
)
file.info("local_embeddings.duckdb")$size
```

## 3. Write embeddings to houba mmatrix

```{r}
system.time(

houba <- writeEmbeddingsHoubaFromDuckDB(dbPath = "local_embeddings.duckdb", 
overwrite = FALSE)
)
houba
file.info("local_embeddings.houba")$size
```

## 4. Run PCA on houba mmatrix

```{r}
system.time(
  pca_res <- houbaPCA("local_embeddings.houba")
)
```

## 5. Get PCA scores

```{r}
pc_scores <- getPcaScores(pca_res)
object.size(pc_scores)
```

## 6. Plot PCA dimensions (PC1 vs PC2, colored by chromosome)

```{r}
plotPcaDims(pc_scores, houba$info, annotation_col = "chrom", dim1 = 1, dim2 = 2)
```

## 7. Annotate variants with coffee consumption GWAS and plot PC1 vs PC2 by gene

```{r}
# Fetch coffee consumption Phenotype data
query <- utils::URLencode("https://rest.ensembl.org/phenotype/term/homo_sapiens/coffee consumption")
resp <- request(query) |>
  req_headers("Content-Type" = "application/json") |>
  req_perform()

coffee_json <- resp_body_json(resp)

# Extract rsid and gene annotation
coffee_anno <- lapply(coffee_json, function(x) {
  if (!is.null(x$Variation)) {
    data.frame(
      rsid = x$Variation,
      gen = if (!is.null(x$attributes$associated_gene)) x$attributes$associated_gene else NA,
      stringsAsFactors = FALSE
    )
  }
}) |> bind_rows()

# Fetch Preeclampsia Phenotype data
query_preeclampsia <- utils::URLencode("https://rest.ensembl.org/phenotype/term/homo_sapiens/preeclampsia")
resp_preeclampsia <- request(query_preeclampsia) |>
  req_headers("Content-Type" = "application/json") |>
  req_perform()
preeclampsia_json <- resp_body_json(resp_preeclampsia)

# Extract rsid and gene annotation for Preeclampsia
preeclampsia_anno <- lapply(preeclampsia_json, function(x) {
  if (!is.null(x$Variation)) {
    data.frame(
      rsid = x$Variation,
      gen = if (!is.null(x$attributes$associated_gene)) x$attributes$associated_gene else NA,
      stringsAsFactors = FALSE
    )
  }
}) |> bind_rows()

# Merge PCA scores and houba info
plotDf <- cbind(pc_scores,houba$info)
names(plotDf)[1:15] <- paste0("PC",1:15)

# Annotate plotDf with coffee and preeclampsia
plotDf <- plotDf |>
  left_join(coffee_anno, by = c("rsid" = "rsid")) |>
  left_join(preeclampsia_anno, by = c("rsid" = "rsid"), suffix = c("_coffee", "_preeclampsia"))

# Subset for coffee and preeclampsia SNPs
coffee_snps <- subset(plotDf, !is.na(gen_coffee))
preeclampsia_snps <- subset(plotDf, !is.na(gen_preeclampsia))

preeclampsia_snps
ggplot() +
  geom_point(data = coffee_snps, aes(x = PC1, y = PC2, color = gen_coffee), shape = 16, size = 2, alpha = 0.7) +
  geom_point(data = preeclampsia_snps, aes(x = PC1, y = PC2, color = gen_preeclampsia), shape = 17, size = 3, alpha = 0.7) +
  labs(title = "PC1 vs PC2: coffee SNPs (circles) and preeclampsia SNPs (triangles)", color = "Gene annotation") +
  theme_minimal()
```

---

- All file sizes are displayed after each step for transparency.
- Remote parquet files are previewed before local processing.
- For large datasets, consider subsampling for faster plotting.
- Use `overwrite = FALSE` for DuckDB and houba steps to avoid slow download and reuse existing files.
